---
title: "Working with Spatial Data"
author: "Zach Schuster"
date: "November 14, 2018"
output: github_document
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(ggmap)
library(plotly)
```

The purpose of this file is to explore and visualize US mass shootings between the last quarter of 2011 and the first day of 2015. The data used is below:

* MassShootingsDataset.csv

In addition, we will call the below API to query coordinates when they are missing but we have the location.

* Texas A&M Geo Services

First read in the data and explore

```{r}
shootings = fread("Mass_Shootings_Dataset.csv", header = TRUE,
                  na.strings = c(""))

# How many na values are there?
vapply(shootings, function(x) sum(is.na(x)),
       FUN.VALUE = integer(1))
```

We can see that we have 20 rows where longitute and latitude are missing, but the location is present. Can we query an API to find these coordinates?

```{r, include=FALSE}
key = readLines("apiKey.txt")
```

To do so, we will subset out the rows we need to fill in and then bind them back to the original data once we have a full data set.

```{r}
to_fill = shootings[is.na(Latitude) | is.na(Longitude)]

# remove to_fill rows from shootings
shootings = shootings[!to_fill, on = names(shootings)]

# There should not be any overlap between to_fill and shootings
intersect(to_fill, shootings)
```

Good! There is no overlap between the two tables. Now we can write code to fill in the missing coordinates

Just a few more steps. It'll be easier when we query to have a city and a state column...

```{r}
cityState = tstrsplit(to_fill$Location, split = ",")
```

We can write a function where we will feed in a vector of cities and a vector of states and get back a list of latitudes and longitudes.

```{r }
getCoords = function(City, State){
  
  #flatten City and State if need be
  City = unlist(City)
  State = unlist(State)
  
  # we need to replace any spaces in the categories with "%20"
  City = gsub("\\s+", "%20", trimws(City))
  State = gsub("\\s+", "%20", trimws(State))
  
  # make sure city and state are of equal length
  stopifnot(length(City) == length(State))
  
  # define constants
  baseQuery = "https://geoservices.tamu.edu/Services/Geocode/WebService/GeocoderWebServiceHttpNonParsed_V04_01.aspx?"
  apiKey = paste("apiKey=", key, sep = "")
  version = "version=4.01"
  format = "format=csv"
  header = "includeHeader=true"
  
  latLon = lapply(seq_along(City), function(i){
  
    # define parameters
    city = paste("city=", City[i], sep = "")
    state = paste("state=", State[i], sep = "")
    
    options = paste(apiKey, version, city, state, format, header, sep = "&")
    
    # query API
    results = fread(paste0(baseQuery, options), verbose = FALSE,
                    showProgress = FALSE)
    
    # return latitude and longitude
    return(c(results$Latitude, results$Longitude))
  })
  
  return(setNames(transpose(latLon), c("Latitude", "Longitude")))

}

# we can transpose the list to make it cleaner
(coords = getCoords(cityState[[1]], cityState[[2]]))
```

Now that we have our latitude and logitudes, we can update the data.table `to_fill` and bind the rows back to the shootings data.table.

```{r}
# fill latitude and logitude
to_fill[, (c("Latitude", "Longitude")) := coords]

# bind to shootings and order by date
shootings = rbind(shootings, to_fill)[order(Date)]
```

I'm not too worried about the one missing summary, but let's take a look at the missing race and gender observations.

```{r}
shootings[is.na(Race) | is.na(Gender), !"Summary"]
```

The las vegas shooting involved a male, so we can fill that in. The Race we will leave as NA for now.

```{r}
shootings[is.na(Gender), Gender := "M"]
```

## Visualizing Shootings

The `ggmap` package will be used for geographical plotting.

Initially, let's look at a simple map of all the shootings in the dataset.

```{r, message=FALSE}
qmplot(Longitude, Latitude,
       data = shootings, maptype = "toner-lite",
       main = "Mass Shootings Dataset")
```

Here we see that there appears to be one incident in Alaska and one in Hawaii, which is why the map is do zoomed out. One point (bottom right of map) looks to be in the ocean. Let's investigate.

```{r}
shootings[Longitude > -25, !"Summary"]
```

There must be an error as Fort Hood definitely does not have coordinates of (0, 0)! This could be a security feature as the API I called returned (0, 0)! We can use another observation since there are multiple entries for Fort Hood

```{r}
Loc = shootings[Location == "Fort Hood, Texas" & Latitude > 0][1]
Loc = Loc[,.(Latitude, Longitude)]

shootings[Longitude > -25, `:=`(Latitude = Loc$Latitude,
                                Longitude = Loc$Longitude)]

shootings[Location == "Fort Hood, Texas"]
```

Great! Our map should look better now.

```{r, message = FALSE}
qmplot(Longitude, Latitude,
       data = shootings, maptype = "toner-lite",
       "Mass Shootings Dataset")
```

It may make our analysis easier if we remove the incidents from AK and HI.

```{r}
shootings = shootings[!(grepl("Hawaii", Location, fixed = TRUE) | 
                          grepl("Alaska", Location, fixed = TRUE))]
```

```{r, message = FALSE}
qmplot(Longitude, Latitude,
       data = shootings, maptype = "toner-lite",
       main = "Mass Shootings Dataset")
```

I think we are now able to get a good picture of the lower 48 US States!

***

From here, how can we get a better idea of the shootings we have seen using color?

We can plot the shootings on a map using color to show gender. 

**Note: We will use the toner-lite map type as it is difficult to visualize points on terrain type maps.**

```{r, message=FALSE}
us = c(left = -125, bottom = 25.25, right = -67, top = 49)
map = get_stamenmap(us, maptype = "toner-lite", zoom = 4)

# recode gender
shootings[, Gender := forcats::fct_collapse(Gender,
                                            Male = c("Male", "M"),
                                            Both = c("M/F", "Male/Female"))]

# create map
ggmap(map) + 
  geom_point(aes(x = Longitude, y = Latitude, colour = Gender),
             data = shootings, alpha = .5) +
  ggtitle("Mass Shootings in the Unites States by Gender")
```

Using the alpha parameter to make points somewhat transparent helps visualize areas of high density. To no surprise, it looks like major metropolitan areas such as Los Angeles, Seattle, Atlanta, and New York are areas where many shootings occur.

```{r}
shootings[, sort(table(Location), decreasing = TRUE)][1:5]
```

After looking at a table, Seattle has the most shootings in the data set. 

We can also see that the shooters are predominantly male.

```{r}
# Percentage of each Gender
round(table(shootings$Gender)/length(shootings$Gender), 2)
```

***

We could do the same analysis for Race

```{r}
shootings[, Race := 
            forcats::fct_collapse(Race,
                                  Asian = c("Asian", "Asian American",
                                            "Asian American/Some other race"),
                                  Black = c("black", "Black",
                                            "Black American or African American",
                                            "Black American or African American/Unknown"),
                                  `Native American` = c("Native American",
                                                        "Native American or Alaska Native"),
                                  Unclear = c("Other", "Some other race",
                                              "Two or more races", "unclear",
                                              "Unknown"),
                                  White = c("white", "White",
                                            "White American or European American",
                                            "White American or European American/Some other Race"))]

# look at a table of race to verify the code worked properly
shootings[, table(Race)]
```

After collapsing levels of Race, we see that Whites have been the criminals far more often than any other race. Let's plot shooting on a map by Race. I will exclude the unclear races.

```{r}
ggmap(map) + 
  geom_point(aes(x = Longitude, y = Latitude, colour = Race),
             data = shootings[Race != "Unclear"],
             alpha = .5) +
  ggtitle("Shootings by Race")

```

I think the results are difficult to interpret if we don't know exactly what we're looking for. As expected, we do see a lot of pink points which correspond to white shooters.

***

We can try to see if there is a visual pattern in shootings that occured by those with mental health issues compared to those without. We will only look at shootings where we know whether or not there were metal health issues.

```{r}
# relevel the mental health issues column
shootings[, Mental_Health_Issues := 
            forcats::fct_collapse(`Mental Health Issues`,
                                  Unknown = c("Unclear", "unknown", "Unknown"))]
shootings[, `Mental Health Issues` := NULL]

# create map
ggmap(map) + 
  geom_point(aes(x = Longitude, y = Latitude, colour = Mental_Health_Issues),
             data = shootings[Mental_Health_Issues != "Unknown"],
             alpha = .5) +
  ggtitle("Shootings by Mental Health Issues")

```

There don't appear to be any major visual patterns. I spent four year in Tampa, FL for my undergraduate degree. It is interesting to see that almost all incidents around that area happened by people without mental health issues.

***

Lastly, we could zoom in on a certain area, Seattle, and overlay a density plot
```{r, message=FALSE}
seattle = c(left = -123.3, bottom = 46.8, right = -121.5, top = 48.8)
map_seattle = get_stamenmap(seattle, zoom = 7)

# create subset for faster plotting
seattleISH = shootings[Longitude < -120 & Latitude > 47]

ggmap(map_seattle) +
  geom_point(aes(x = Longitude, y = Latitude), data = seattleISH, col = "red") +
  geom_density2d(aes(x = Longitude, y = Latitude), data = seattleISH,
                 col = "blue") + 
  ggtitle("Density of Seattle Shootings")
```


Although the graphic is not perfect due to the sparsity of data outside of Seattle, we can see how much denser Seattle is than any of the surrounding areas. 

## Plotly Interactive Maps

Example code is below. The map is not shown as it can only be used in an html format (which this .md file is not)

```{r, message=FALSE, eval = FALSE}

# geo styling
g <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showland = TRUE,
  landcolor = toRGB("gray95"),
  subunitcolor = toRGB("gray85"),
  countrycolor = toRGB("gray85"),
  countrywidth = 0.5,
  subunitwidth = 0.5
)

# make plot and hover information
p <- plot_geo(shootings, lat = ~Latitude, lon = ~Longitude) %>%
  add_markers(
    text = ~paste(Title, Location, Date, paste("Deaths:", Fatalities),
                  sep = "<br />"),
    hoverinfo = "text"
  ) %>%
  layout(
    title = 'US Mass Shootings<br />(Hover for info)', geo = g
  )

# show plot
p
```


## Future Ideas

* Use plotly to create interactive maps. We could show information on each shooting when a cursor hovers over it and much more given the time. 
* Create a Google Maps API key to overlay points on Google Maps, which would show more detail. 

*Some of the ideas and methods from this analysis were adapted from the [ggmap github page](https://github.com/dkahle/ggmap) and [this plotly tutorial](https://plot.ly/r/scatter-plots-on-maps/). Thank you for great examples!*

**This short analysis has been an exercise in plotting spatial data. Please let me know if you have any suggestions. I'm always interested in better and more efficient programming practices!**

